D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-01 ootavacuous.tex
Global Optimization...: Loose C++ and avoid OOTA
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which \co{thread1()} loads the value of \co{x} and stores it in
\co{y} while \co{thread2()} does the reverse.
A globally optimizing compiler given that program might
transform it to the following before translating it into machine code,
if the compiler is sufficiently perverse:
\begin{quote}
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   int r1 = 42;
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   int r2 = 42;
13   x.store(r2, memory_order_relaxed);
14 }
\end{verbatim}
\end{quote}
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

The only justification a compiler could have for generating output like
this is that it knows exactly what accesses will be performed by both
threads, and therefore it knows that it will not violate the loose C++
memory model by assuming each thread's load reads from the other's
store.
A similar justification can underlie the reasoning in
section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations};
in principle an analysis of the complete program could lead a compiler
to conclude that \co{y} will always be equal to \co{z} whenever a
particular \co{y - z} expression is evaluated, allowing the compiler
to replace the expression with a constant \co{0}.

>>>>

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which \co{thread1()} loads the value of \co{x} and stores it in
\co{y} while \co{thread2()} does the reverse.
A globally optimizing loose C++ compiler given that program might
transform it to the following before translating it into machine code,
if the compiler is sufficiently perverse:
\begin{quote}
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   int r1 = 42;
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   int r2 = 42;
13   x.store(r2, memory_order_relaxed);
14 }
\end{verbatim}
\end{quote}
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

The only justification a compiler could have for generating output like
this is that it knows exactly what accesses will be performed by both
threads, and therefore it knows that it will not violate the loose C++
memory model by assuming each thread's load reads from the other's
store.\footnote{
	Such a compiler could presumably instead choose to avoid this
	OOTA cycle, for example, by retaining the loads from \co{x}
	and \co{y}.}
A similar justification can underlie the reasoning in
section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations};
in principle an analysis of the complete program could lead a compiler
to conclude that \co{y} will always be equal to \co{z} whenever a
particular \co{y - z} expression is evaluated, allowing the compiler
to replace the expression with a constant \co{0}.
